Describing the scale and composition of calls for police service: a replication and extension using open data

### 2022-02-15 paper.Rmd
update paper

---
title: 'Describing the scale and composition of calls for police service: a replication and extension using open data'
author: Samuel Langton\dag, Stijn Ruiter\dag\*, Tim Verlaan\dag
bibliography: refs.bib
output:
  pdf_document: 
    number_sections: true
header-includes:
  - \usepackage{lscape}
  - \usepackage{lineno}
  - \linenumbers
csl: apa.csl
abstract: "This paper describes the scale and composition of emergency demand for police services in Detroit, United States. The contribution is made in replication and extension of analyses reported by Ratcliffe (2021) and Lum et al. (2021). Findings indicate that police spend a considerable proportion of time performing a social service function. Just 47% of total deployed time responding to 911 calls is consumed by crime incidents. The remainder is spent on community (6%), health (7%), proactive (5%), quality of life (19%), and traffic (13%) duties. A small number of calls consume a disproportionately large amount of police officer time. Emergency demand is concentrated in time and space, and can differ between types of demand. Findings further highlight the potential implications of radically reforming police forces in the United States. The data and code used here are openly available for reproduction, re-use and scrutiny."
---

@@

The 'time spent' measure used was the _time on scene_ (i.e., excluding the response time). The authors considered this to be preferable, since low priority calls may have a non-urgent (slow) response due to the nature of the call, and officers may be reassigned to other duties while on an assignment. Lum and colleagues calculated _time on scene_ as the time from the first officer's arrival to when the call was reported as resolved. In this way, the calculation did not account for the additional time consumed by multi-officer responses (e.g., back-up). This was justified for a number of reasons, among them, avoiding overinflation of time due to officers 'flocking' to non-serious incidents, and limitations with the CAD data. A proportional breakdown of the disposition (assist only, report written or arrest citation) was provided for each of the fourteen call types. No spatial or temporal findings were reported.

@@

For each incident, the police response time and time on the scene are reported. The former measures the response time from dispatch to when the first officer arrives on-scene.^[In the raw data, response time is referred to as _traveltime_. This figure does not include the intake or dispatch time which we did not deem to be deployment.] The latter mimics the measure used by @lum2021can, namely, the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. Here, we refer to the sum of the response time and time on scene as the _total deployed time_. Calls which had a response time of negative values or missing were dropped from analyses. Calls which had a _time on scene_ of zero minutes, negative values or missing were also excluded. The data permit a distinction between the time spent with and without the response time. As noted, @lum2021can considered the _time on scene_ measure to be preferable. Here, for completeness, we report times with and without the response time.

@@

The frequency counts and proportional breakdowns for each demand type classification are reported in Table 1. This demonstrates that in Detroit, a considerable proportion of citizen calls for service, and in turn, time spent by officers, is committed to a diverse array of (often non-criminal) issues. Just 44% of emergency calls for service related to crime incidents. These crime incidents consume 51% of time on scene throughout the year. The remainder of total time on scene is spent dealing with traffic duties (16%), quality of life incidents (15%), public and mental health (7%), community issues (5%), and proactive policing (4%). The differences in the proportional breakdown between time on scene and total deployment time are marginal.

@@

On this point, this study raises useful discussion points around the different measures of officer time. First, it is clear that findings on the scale and composition of police demand are sensitive, albeit only slightly, to whether measures of officer time include the response time. The degree to which the two estimates differ varies by demand classification, likely as a result of the different call priorities, and in turn, the rapidness of the response times. That said, this claim would need verifying through detailed examination. On this point, second, we would reiterate that the 'time on scene' measure reported here, in alignment with @lum2021can, refers to the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. This differs to that of @ratcliffe2021policing which accounted the number of officers attending the scene. @lum2021can note that including back-up officers in the time calculation might introduces complications. For instance, non-serious incidents might receive back-up unnecessarily, particularly during periods of low-demand, which would inflate the time consumed and overstate the resources required to resolve such an incident. Although the overall composition of police demand is comparable between Philadelphia [@ratcliffe2021policing] and Detroit, the true impact of these measures remains unclear. We propose that future research investigates these distinctions. This would require CAD (or equivalent) data, timestamped at each stage the response (call-handler dispatch, arrival on scene, scene departure), for individual police units nested within calls. This level of granularity would permit the calculation and comparison of multiple measures of officer time, as well as an investigation into the extent to which back-ups and/or 'flocking' might influence such measures.

>>>>

---
title: 'Describing the scale and composition of calls for police service: a replication and extension using open data'
author: Samuel Langton\dag, Stijn Ruiter\dag\*, Tim Verlaan\dag
bibliography: refs.bib
output:
  pdf_document: 
    number_sections: true
header-includes:
  - \usepackage{lscape}
  - \usepackage{lineno}
  - \linenumbers
csl: apa.csl
abstract: "This paper describes the scale and composition of emergency demand for police services in Detroit, United States. The contribution is made in replication and extension of analyses reported by Ratcliffe (2021) and Lum et al. (2021). Findings indicate that police spend a considerable proportion of time performing a social service function. Just 51% of total deployed time responding to 911 calls is consumed by crime incidents. The remainder is spent on community (5%), health (7%), proactive (4%), quality of life (16%), and traffic (15%) duties. A small number of calls consume a disproportionately large amount of police officer time. Emergency demand is concentrated in time and space, and can differ between types of demand. Findings further highlight the potential implications of radically reforming police forces in the United States. The data and code used here are openly available for reproduction, re-use and scrutiny."
---

@@

The 'time spent' measure used was the _time on scene_ (i.e., excluding the response time). The authors considered this to be preferable, since low priority calls may have a non-urgent (slow) response due to the nature of the call, and officers may be reassigned to other duties while on an assignment. Lum and colleagues calculated _time on scene_ as the time from the first officer's arrival to when the call was reported as resolved. In this way, the calculation did not account for the additional time consumed by multi-officer responses (e.g., back-up). This was justified for a number of reasons, among them, avoiding the over-inflation of time estimates due to officers 'flocking' to non-serious incidents, and limitations with the CAD data. A proportional breakdown of the disposition (assist only, report written or arrest citation) was provided for each of the fourteen call types. No spatial or temporal findings were reported.

@@

For each incident, the police response time and time on the scene are reported. The former measures the time from dispatch to when the first officer arrives on-scene.^[In the raw data, the time between dispatch and first on-scene arrival is referred to as _travel time_. The _intake_ and _dispatch time_ are reported, however, we did not consider these measures to reflect deployment time. We return to this point in the discussion.] The latter mimics the measure used by @lum2021can, namely, the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. Here, the data permit a distinction between the time spent on calls with and without the response time. We refer to the sum of the response time and time on scene as the _total deployed time_. Calls which had a response time of negative values or missing were dropped from analyses. Calls which had a _time on scene_ of zero minutes, negative values or missing were also excluded. As noted, @lum2021can considered the _time on scene_ measure to be preferable. For completeness, in this study we report times with and without the response time -- defined as travel time.

@@

The frequency counts and proportional breakdowns for each demand type classification are reported in Table 1. This demonstrates that in Detroit, a considerable proportion of citizen calls for service, and in turn, time spent by officers, is committed to a diverse array of (often non-criminal) issues. Just 44% of emergency calls for service related to crime incidents. These crime incidents consume 51% of time on scene throughout the year. The remainder of time on scene is spent dealing with traffic duties (16%), quality of life incidents (15%), public and mental health (7%), community issues (5%), and proactive policing (4%). The differences in the proportional breakdown between time on scene and total deployment time, which includes response time, are marginal. This indicates that the response time plays only a minor role determining the scale and composition of time consumed.

@@

On this point, this study raises useful discussion points around the different measures of officer time. First, it is clear that findings on the scale and composition of police demand are only marginally sensitive to whether measures of officer time include the response time -- defined as the time from dispatch to when the first officer arrives on-scene. The overall findings are almost identical between the two measures. Here, we would highlight that our response time measure did not include dispatch time, but rather, travel time in isolation. Including dispatch time would increase the gap between the two measures: low-priority calls will be held in the dispatch queue for longer, inflating the overall response time, while high-priority calls will be dispatched immediately. Since certain calls will have higher priority responses (e.g., violent crime) this gap is likely to differ between demand classifications. We excluded dispatch time on the basis that it does not capture deployment, but a specific investigation into its impact would be straightforward using the open materials provided.

Second, and relatedly, we would reiterate that the 'time on scene' measure reported here, in alignment with @lum2021can, refers to the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. This differs to that of @ratcliffe2021policing which accounted for the number of officers attending the scene. @lum2021can note that including back-up officers in the time calculation might introduces complications. For instance, non-serious incidents might receive back-up unnecessarily, particularly during periods of low-demand, which would inflate the time consumed and overstate the resources required to resolve such an incident. Although the overall composition of police demand is comparable between Philadelphia [@ratcliffe2021policing] and Detroit, the true impact of these measures remains unclear. We propose that future research investigates these distinctions. This would require CAD (or equivalent) data, timestamped at each stage of the response (dispatch, scene arrival, scene departure), for individual police units nested within calls. This level of granularity would permit the calculation and comparison of multiple measures of officer time, as well as an investigation into the extent to which back-ups and/or 'flocking' might influence such measures.
